
\subsection{Old version}

It turns out that we cannot use linearisability testing directly with the
specification object~|TwoLinSpec| from Section~\ref{sec:twoStepLinSpec},
because it gives false errors caused by delays in writing to the log.  We
describe in more detail how such testing might be done, and then explain the
cause of the false errors.  A thread that performs the concrete
operation~$\sm{op}_1(x_1)$: (1)~writes $\call.\sm{op}_1^i(x_1)$ into the log;
(2)~performs $\sm{op}_1(x_1)$ on the synchonisation object, obtaining
result~$y_1$; (3)~writes $\return.\sm{op}_1^i \:: ()$,\,
$\call.\overline{\sm{op}}_1^i()$ and $\return.\overline{\sm{op}}_1^i \:: y_1$
into the log.  A thread that performs operation~|op|\s2 acts as for standard
linearisation testing.  Once all threads have finished, we could use the
existing algorithms for testing whether the history is linearisable with
respect to |TwoStepLinSpec|.

This approach does not work, because it gives false errors.  For example, the
timeline below depicts a log that could be obtained from a correct synchronous
channel using the above approach, where we treat |send| as |op|\s1.
%
\begin{center}
\unScalaMid
\begin{tikzpicture}
\draw[|-|] (0,0) -- node[above] {  $\sm{send}^1(5)$ } (5.5,0);
\draw[|-|] (6,0) -- node[above] { $\overline{\sm{send}}^1 \:: ()$ } (8,0);
\draw[|-|] (0.5,-1) -- node[above] { $\sm{receive}^2 \:: 5$} (1.5,-1) ;
\draw[|-|] (2.5,-2) -- node[above] {  $\sm{send}^3(6)$ } (4,-2);
\draw[|-|] (5,-2) -- node[above] { $\overline{\sm{send}}^3 \:: ()$ } (6.5,-2);
\draw[|-|] (2.0,-3)  -- node[above] { $\sm{receive}^4 \:: 6$} (4.5,-3);
\end{tikzpicture}
\scalaMid
\end{center}
%
Here, the invocations in the top two rows synchronise to transmit~5, and then
the invocations in the bottom two rows synchronise to transmit~6.  However,
the thread for the top row is slow to write its last three events into the
log.  The above history is not linearisable with respect to |TwoStepLinSpec|:
it is clear that $\sm{send}^1(5)$ and $\sm{receive}^2 \:: 5$ would need to be
linearised first; but this would require $\overline{\sm{send}}^1$ to be
linearised before $\sm{send}^3(6)$, which is inconsistent with the history.
Hence the approach would generate a false error.

Instead we use a technique that is robust against delays in logging.  We start
by considering the binary heterogeneous case, i.e.~each synchronisation is
between precisely two threads, which have called different operations.  We
assume that each thread has an identity in some range $\range 0
{\sm{NumThreads}}$.  We arrange for this identity to be included in the
$\call$ events written to the log for operations~|op|\s1 and
$\overline{\sm{op}}_1$, but otherwise threads act as above; in particular, for
each thread, calls to~|op|\s1 and $\overline{\sm{op}}_1$ alternate.

We then test whether the history is linearisable with respect to the
specification object below.  This object requires that corresponding
invocations of~|op|\s1 and~|op|\s2 are linearised consecutively: it encodes
the automaton on the right.  However, it allows
the corresponding $\overline{\sm{op}}_1$ to be linearised later (but before
the next operation invocation by the same thread).  It uses an array
|returns|, indexed by thread identities, to record values that should be
returned by a $\overline{\sm{op}}_1$ operation.
%
\begin{trivlist}
\item[]
\begin{minipage}{92mm}
\begin{scala}
type ThreadID = Int               // Thread identifiers
val NumThreads: ThreadID = ... // Number of threads
trait State
case class Zero extends State
case class One(t: ThreadID, x£\s1£: A£\s1£) extends State
\end{scala}
\end{minipage}
%%%%%
\hfill 
%
\begin{minipage}{37.8mm}
\begin{tikzpicture}[>= angle 60, xscale = 0.9, yscale = 0.44]
\draw (0,0) node[draw] (zero) {$\sm{Zero}$};
\draw[->] (zero) ++ (-1.5, 0) -- (zero);
%
\draw (0,-4) node[draw] (one) {$\sm{One}(\sm t, \sm{x}_1)$};
\draw[->] (zero) .. controls ++(0.3,-2) .. 
  node[right] {$\sm{op}_1(\sm t, \sm{x}_1)$} (one); 
%
\draw[->] (one) .. controls ++(-0.3,2) .. 
  node[left] {$\sm{op}_2(\sm x_2)$} (zero);
\end{tikzpicture}%
\end{minipage}%
%%%%%
\begin{scala}
object TwoStepDelayedLinSpec{
  private var state: State = Zero
  private val returns = new Array[Option[B£\s1£]](NumThreads)
  for(t <- 0 until NumThreads) returns(t) = None
  def op£\s1£(t: ThreadID, x£\s1£: A£\s1£): Unit = {
    require(state.isInstanceOf[Zero]); state = One(t, x£\s1£); ()
  }
  def op£\s2£(x£\s2£: A£\s2£): B£\s2£ = {
    require(state.isInstanceOf[One]); val One(t, x£\s1£) = state
    val (y£\s1£, y£\s2£) = SyncSpec.sync(x£\s1£, x£\s2£); returns(t) = Some(y£\s1£); state = Zero; y£\s2£
  }
  def £$\overline{\sm{op}}_1$£(t: ThreadID): B£\s1£ = {
    require(returns(t).isInstanceOf[Some]); val Some(y£\s1£) = returns(t)
    returns(t) = None; y£\s1£
  }
}
\end{scala}
\end{trivlist}

The following lemma identifies important properties of
|Two|\-|Step|\-|Delayed|\-|LinSpec|.  It follows immediately from the
definition.
%
\begin{lemma}
\label{lem:TwoStepDelayedLinSpec}
Within any legal history of |TwoStepDelayedLinSpec|, events $\op_1$ and
$\op_2$ alternate.  Let $\op_1^{i_1}(t,x_1) \:: ()$ and $\op_2^{i_2}(x_2) \::
y_2$ be a consecutive pair of such events.  Then |op|\s2 makes a call
$\sm{SyncSpec.sync}(x_1, x_2)$ obtaining result $(y_1,y_2)$.  Under the
assumptions about threads within the test, the next event for thread~$t$ will
be $\overline\op_1^{i_1}(t) \:: y_1$; and this will be later in the history
than $\op_2^{i_2}(x_2) \:: y_2$.  Further, the corresponding history of
events $\sm{sync}^{i_1,i_2}(x_1,x_2) \:: (y_1,y_2)$ is a legal history of
|SyncSpec|.

Conversely, each history with events ordered in this way will be a legal
history of |TwoStepDelayedLinSpec| if  the corresponding history
of events $\sm{sync}^{i_1,i_2}(x_1,x_2) \:: (y_1,y_2)$ is a legal history of
|SyncSpec|.
\end{lemma}

%%%%%

In order to argue for correctness, we need to distinguish between:
%
\begin{itemize}
\item the \emph{invocation history}, in terms of actual calls and returns of
  |op|\s1 and |op|\s2; and

\item the corresponding \emph{log history}, which might contain delays.
\end{itemize}
%
For clarity, we annotate events with ``$inv$'' or~``$log$''.  The invocation
history uses events of the form $\call^{inv}.\sm{op}_1^{i_1}(x_1)$,\,
$\return^{inv}.\sm{op}_1^{i_1} \:: y_1$,\,
$\call^{inv}.\sm{op}_2^{i_2}(x_2)$,\, and $\return^{inv}.\sm{op}_2^{i_2} \::
y_2$.  The log history uses events of the form
$\call^{log}.\sm{op}_1^{i_1}(t,x_1)$ (note the additional thread identity
parameter), $\return^{log}.\sm{op}_1^{i_1} \:: ()$ (note the unit return
value), $\call^{log}.\overline{\sm{op}}_1^{i_1}(t)$,\,
$\return^{log}.\overline{\sm{op}}_1^{i_1} \:: y_1$ (note the transferred
return value), $\call^{log}.\sm{op}_2^{i_2}(x_2)$, and
$\return^{log}.\sm{op}_2^{i_2} \:: y_2$

We can consider the interleaving of the invocation and log histories,
following real-time order.  In the interleaving, for |op|\s1 and |op|\s2,\,
$\call^{log}$ events will be earlier than the corresponding $\call^{inv}$
events; and $\return^{log}$ events will be later than the corresponding
$\return^{inv}$ events.  However, the two histories agree on the relative
order of the |op|\s1 and |op|\s2 events of each individual thread.

The proposition below shows that this testing method does not generate any false
errors. 
%
\begin{prop}
Suppose synchronisation object |SyncObj| is linearisable with respect
to~|SyncSpec|.  Then each complete log history~$h_l$ of |SyncObj| is
linearisable with respect to |TwoStepDelayedLinSpec|.
\end{prop}

%%%%%

\begin{proof}
Consider a complete log history~$h_l$ of~|SyncObj|, and a corresponding
invocation history~$h$; and consider their real-time interleaving.  By
assumption, there is a legal history~$h_s$ of |SyncSpec| such that $h$
and~$h_s$ are synchronisation compatible.  Thus $h_s$ may be interleaved with
the interleaving of~$h$ and~$h_l$, so that each $\sm{sync}$ event from~$h_s$
is between the corresponding $\call^{inv}$ and $\return^{inv}$ events
from~$h$, and hence also between the corresponding $\call^{log}$ and
$\return^{log}$ events from~$h_l$.

We build an interleaving of $h$, $h_l$ and a legal history $h_s'$ of
|Two|\-|Step|\-|Delayed|\-|LinSpec| from the interleaving of~$h$, $h_l$
and~$h_s$, as follows.
%
\begin{enumerate}
\item We replace each $\sm{sync}^{i_1,i_2}(x_1,x_2) \:: (y_1,y_2)$ by
  the two (consecutive) events $\sm{op}_1^{i_1}(t,x_1) \:: ()$ and
  $\sm{op}_2^{i_2}(x_2) \:: y_2$, where $t$ is the identity of the thread that
  makes the corresponding call of |op|\s1 in~$h_l$.  

\item We insert an event $\overline{\sm{op}}_1^{i_1}(t) \:: y_1$ between every
  $\call^{log}.\overline{\sm{op}}_1^{i_1}(t)$ and
  $\return^{log}.\overline{\sm{op}}_1^{i_1} \:: y_1$ (from~$h_l$).
  %% , but not between any pair of $\sm{op}_1$ and $\sm{op}_2$ events from the
  %% previous stage.  \framebox{???}
\end{enumerate}
%
Note that each inserted event is between the corresponding $\call^{log}$ and
$\return^{log}$ events, by construction.  Let $h_s'$ be these inserted events;
we show that $h_s'$ is a legal history of |Two|\-|Step|\-|Delayed|\-|LinSpec|.
\begin{itemize}
\item The events inserted in step~1 alternate between |op|\s1 and~|op|\s2, as
  required by |Two|\-|Step|\-|Delayed|\-|LinSpec|.  Further, they are in the
  same order, and have the same values for~$x_1$, $x_2$, $y_2$, $i_1$
  and~$i_2$ as the corresponding |sync| events from~$h_s$.  Hence each
  inserted |op|\s2 event has the return value~$y_2$ as required by
  |Two|\-|Step|\-|Delayed|\-|LinSpec|.  Further, the value for~$t$ matches
  that in the corresponding $\call^{log}.\sm{op}_1^{i_1}(t,x_1)$ event; so
  the value $y_1$ written into $\sm{returns}(t)$ (by~|op|\s2 in
  |Two|\-|Step|\-|Delayed|\-|LinSpec|) matches the value returned by the
  corresponding call to~|sync|.

\item For the events from step~2, because of the way logging is done, each
  value of~$y_1$ returned by $\overline{\sm{op}}_1$ must match the value
  returned by the previous invocation of~|op|\s1 on the synchronisation object
  by thread~$t$.  Since~$h$ is synchronisation linearisable, this $y_1$ must
  match the value returned by the corresponding call of~|sync|.  And this
  matches the last value written into $\sm{returns}(t)$ (by the previous
  item), as required by |Two|\-|Step|\-|Delayed|\-|LinSpec|.
\end{itemize}
%
This demonstrates that $h_l$ is linearisable with respect to
|Two|\-|Step|\-|Delayed|\-|LinSpec|.
\end{proof}

%%%%%%%%%%

In theory, the delays in logging can mean that an invocation history that is
not synchronisation linearisable is transformed into a log history that is
linearisable with respect to |Two|\-|Step|\-|Delayed|\-|LinSpec| (although
this seems unlikely).  
%
We show that if the invocation history is not synchronisation linearisable
with respect to |SyncSpec|,
then there is a corresponding log history that is not linearisable with
respect to |Two|\-|Step|\-|Delayed|\-|LinSpec|.
%
\begin{prop}
\label{prop:TwoStepDelayedLinSpec-complete}
Let $h$ be a history of~|SyncObj| that is not synchronisation linearisable
with respect to |SyncObj|.  Then there is a corresponding log history~$h_l$
that is not linearisable with respect to |Two|\-|Step|\-|Delayed|\-|LinSpec|.
\end{prop}

%%%%%

\begin{proof}
Let $h$ be as in the statement of the proposition.  We build a corresponding
log history~$h_l$, and interleave it with~$h$ as follows.  The construction is
illustrated in Figure~\ref{fig:TwoStepDelayedLinSpec-complete}.
%
\begin{itemize}
\item We add an event $\call^{log}.\op_1^{i_1}(t,x_1)$ immediately before
  each $\call^{inv}.\op_1^{i_1}(x_1)$, where $t$ is the identity of the
  thread making the call.

\item We add events $\return^{log}.\op_1^{i_1} \:: ()$,\,
  $\call^{log}.\overline{\op}_1^{i_1}(t)$, and
  $\return^{log}.\overline{\op}_1^{i_1} \:: y_1$ immediately after each
  $\return^{inv}.\op_1^{i_1} \:: y_1$, where again $t$ is the identity of the
  relevant thread.

\item We add an event $\call^{log}.\op_2^{i_2}(x_2)$ immediately before
  each $\call^{inv}.\op_2^{i_2}(x_2)$.

\item We add an event $\return^{log}.\op_2^{i_2} \:: y_2$ immediately after
  each  $\return^{inv}.\op_2^{i_2} \:: y_2$.
\end{itemize}
%
%%%%%
%
\begin{figure}
\def\sep{1.5}
\begin{center}
%% \begin{tikzpicture}
%% \draw(0,0) node (h) {$h$:}; 
%% \draw(0,-\sep) node (h_l) {$h_l$:};
%% \draw(0,-\sep-\sep) node (h-s) {$h_s$:};
%% %% call.op_1
%% \draw (h)++(2,-0.2) node (call-inv-1) {\footnotesize $\call^i.\op_1$};
%% \draw[fill] (call-inv-1)++(0,0.4) circle (2pt); 
%% \draw (h_l)++(1.7,-0.2)node (call-log-1) {\footnotesize $\call^l.\op_1$};
%% \draw[fill] (call-log-1)++(0,0.4) circle (2pt); 
%% %% call.op_2 
%% \draw (call-inv-1)++(2,0) node (call-inv-2) {\footnotesize $\call^i.\op_2$};
%% \draw[fill] (call-inv-2)++(0,0.4) circle (2pt); 
%% \draw (call-log-1)++(2,0) node (call-log-2) {\footnotesize $\call^l.\op_2$};
%% \draw[fill] (call-log-2)++(0,0.4) circle (2pt);
%% %% return.op_1
%% \draw (call-inv-2)++(3,0) node (ret-inv-1) {\footnotesize $\return^i.\op_1$};
%% \draw[fill] (ret-inv-1)++(0,0.4) circle (2pt); 
%% %
%% \draw (call-log-2)++(3.6,0) node (ret-log-1) {\footnotesize $\return^l.\op_1$};
%% \draw[fill] (ret-log-1)++(0,0.4) circle (2pt); 
%% %
%% \draw (ret-log-1)++(0.7,0.8) node (call-log-1-bar) 
%%                                  {\footnotesize $\call^l.\overline\op_1$};
%% \draw[fill] (call-log-1-bar)++(0,-0.4) circle (2pt); 
%% %
%% \draw (ret-log-1)++(1.8,0) node (ret-log-1-bar) 
%%                                {\footnotesize $\return^l.\overline\op_1$};
%% \draw[fill] (ret-log-1-bar)++(0,0.4) circle (2pt); 
%% %% return.op_2
%% \draw (ret-inv-1)++(4,0) node (ret-inv-2) {\footnotesize $\return^i.\op_2$};
%% \draw[fill] (ret-inv-2)++(0,0.4) circle (2pt); 
%% %
%% \draw (ret-inv-2)++(0.3,-\sep) node (ret-log-2)
%%    {\footnotesize $\return^l.\op_2$};
%% \draw[fill] (ret-log-2)++(0,0.4) circle (2pt); 
%% %% op_1
%% \draw (call-inv-1)++(1,-\sep-\sep) node (op-1) {\footnotesize $\op_1$};
%% \draw[fill] (op-1)++(0,0.4) circle (2pt); 
%% %% op_2
%% \draw (call-inv-2)++(1,-\sep-\sep) node (op-2) {\footnotesize $\op_2$};
%% \draw[fill] (op-2)++(0,0.4) circle (2pt); 
%% %% \overline{op}_1
%% \draw (call-log-1-bar)++(0.5,-\sep-0.8)  node (op-1-bar)
%%   {\footnotesize $\overline\op_1$};
%% \draw[fill] (op-1-bar)++(0,0.4) circle (2pt); 
%% \end{tikzpicture}
%%%%%%
\unScalaMid
\begin{tikzpicture}
\draw(0,0) node (h) {$h$:}; 
\draw(0,-\sep) node (h_l) {$h_l$:};
\draw(0,-\sep-\sep) node (h-s) {$h_s$:};
% h
\draw (h)++(2,0) node (call-inv-1){}; 
\draw (call-inv-1)++(2,-0.2) node (call-inv-2){};
\draw (call-inv-2)++(3,0.2) node (ret-inv-1){};
\draw (ret-inv-1)++(2.5,-0.2) node (ret-inv-2){};
\draw[|-|] (call-inv-1) -- node[above]{ $\op_1^{inv}$ } (ret-inv-1);
\draw[|-|] (call-inv-2) -- node[below]{ $\op_2^{inv}$ } (ret-inv-2);
% h_l
\draw (h_l)++(1.7,-0) node (call-log-1) {};
\draw (call-log-1)++(2,-0.2) node (call-log-2) {};
\draw (call-log-2)++(3.6,0.2) node (ret-log-1) {};
\draw (ret-log-1)++(0.2,0) node (call-log-1-bar) {};
\draw (call-log-1-bar)++(1.0,0) node (ret-log-1-bar) {};
\draw (call-log-2)++(6,0) node (ret-log-2) {};
\draw[|-|] (call-log-1) -- node[above]{ $\op_1^{log}$ } (ret-log-1);
\draw[|-|] (call-log-1-bar) -- node[above]{ $\overline\op_1^{log}$ } (ret-log-1-bar);
\draw[|-|] (call-log-2) -- node[below]{ $\op_2^{log}$ } (ret-log-2);
%% op_1
\draw (call-inv-1)++(1,-\sep-\sep) node (op-1) {\footnotesize $\op_1$};
\draw[fill] (op-1)++(0,0.4) circle (2pt); 
%% op_2
\draw (call-inv-2)++(1,-\sep-\sep+0.2) node (op-2) {\footnotesize $\op_2$};
\draw[fill] (op-2)++(0,0.4) circle (2pt); 
%% \overline{op}_1
\draw (call-log-1-bar)++(0.5,-\sep)  node (op-1-bar)
  {\footnotesize $\overline\op_1$};
\draw[fill] (op-1-bar)++(0,0.4) circle (2pt); 
\end{tikzpicture}
\scalaMid
\end{center}
%%%%%
\caption{Representation of the construction in the proof of
  Proposition~\ref{prop:TwoStepDelayedLinSpec-complete}.  
  %% The interleaving of the histories is to be read from left to right.  The
  $\call$s of $\op_1$ and~$\op_2$ might be in either order, as might their
  $\return$s.  The proof shows that both $\call$s precede both $\return$s.}
  %%  We abbreviate ``$log$'' and ``$inv$'' to~``$l$'' and~``$i$'',
  %% respectively.}
\label{fig:TwoStepDelayedLinSpec-complete}
\end{figure}
%
%%%%%
%
Let $h_l$ be these inserted events.  We need to show that $h_l$ is not
linearisable with respect to |Two|\-|Step|\-|Delayed|\-|LinSpec|.  We argue by
contradiction: we suppose that $h_l$ \emph{is} linearisable, and deduce that
$h$ is synchronisation linearisable with respect to |SyncSpec|.

So suppose $h_l$ is linearisable, and let $h_s$ be the corresponding legal
history of~|Two|\-|Step|\-|Delayed|\-|LinSpec|.  
%
Let $\op_1^{i_1}(t,x_1)\::y_1$ and $\op_2^{i_2}(x_2)\::y_2$ be two consecutive
events in~$h_s$.  Then:
%
\begin{enumerate}
\item In $h_l$, the event $\call^{log}.\op_1^{i_1}(t,x_1)$ is before
  $\return^{log}.\op_2^{i_2} \:: y_2$ (to satisfy linearisation).  So in~$h$,
  the event $\call^{inv}.\op_1^{i_1}(t,x_1)$ is before
  $\return^{inv}.\op_2^{i_2} \:: y_2$, by the way we have constructed~$h_l$.

\item In $h_s$, the event $\op_2^{i_2}(x_2)\::y_2$ must be before
  $\overline\op_1^{i_1}(t)\::y_1$ (by Lemma~\ref{lem:TwoStepDelayedLinSpec}).
  So in~$h_l$, the event $\call^{log}.\op_2^{i_2}(x_2)$ must be before
  $\return^{log}.\overline{\op}_1^{i_1} \:: y_1$ (to satisfy linearisation).
  Hence, by the way we have constructed~$h_l$, the event
  $\call^{inv}.\op_2^{i_2}(x_2)$ is before $\return^{inv}.\op_1^{i_1} \::
  y_1$.
\end{enumerate}
%
To summarise the above two points: the invocations of $\op_1^{i_1}$ and
$op_2^{i_2}$ overlap in~$h$. 

Further, consider two such pairs of invocations, $(\op_1^{i_1}, op_2^{i_2})$
and $(\op_1^{i_1'}, op_2^{i_2'})$.  If the overlap in~$h$ of the first pair is
before the overlap of the second pair, then the same is true in~$h_l$ (again
by the way we have constructed~$h_l$), so the corresponding events in~$h_s$
must be ordered in the same way.  Informally, the order of the overlaps is
consistent with~$h_s$. 

Let $h_s'$ be the legal history of |SyncSpec| corresponding to~$h_s$, implied
by Lemma~\ref{lem:TwoStepDelayedLinSpec}.  Given the interleaving of~$h$
and~$h_l$, and the interleaving of~$h_l$ and~$h_s$, we build an interleaving
of $h$ and~$h_s'$.  We insert each event $\sm{sync}^{i_1,i_2}(x_1,x_2)\::
(y_1,y_2)$ from~$h_s'$ during the overlap between the corresponding
invocations in~$h$, and respecting the order of~$h_s$ and~$h_s'$.  This is
possible by the above results.  

%% We insert each event $\sm{sync}^{i_1,i_2}(x_1,x_2)\::
%% (y_1,y_2)$ from~$h_s'$ after both $\call^{inv}.\op_1^{i_1}(t,x_1)$ and
%% $\call^{inv}.\op_2^{i_2}(x_2)$, but before both $\return^{inv}.\op_1^{i_1} \::
%% y_1$ and $\return^{inv}.\op_2^{i_2} \:: y_2$.  Note that this is possible,
%% because we have shown that $\call^{inv}.\op_1^{i_1}(t,x_1)$ is before
%% $\return^{inv}.\op_2^{i_2} \:: y_2$ (item~1 above), and
%% $\call^{inv}.\op_2^{i_2}(x_2)$ is before $\return^{inv}.\op_1^{i_1} \:: y_1$
%% (item~2 above).  Clearly this interleaving preserves the order of~$h_s'$.

Hence we have shown that $h$ is synchronisation compatible with~$h_s'$, and so
is linearisable with respect to |SyncSpec|, as required.
\end{proof}

%%%%%

The above proposition included the possibility of an invocation history~$h$ of
|SyncObj| not being synchronisation linearisable, but a corresponding log
history~$h_l$ being linearisable with respect to |TwoStepDelayedLinSpec|,
i.e.~the incorrect history not being detected.  Let us examine this
possibility in more detail.  So suppose $h$ is not synchronistion
linearisable, but $h_l$ is linearisable.  Let $h_s$ be the corresponding legal
history of |TwoStepDelayedLinSpec|, and let $h_s'$ be the corresponding legal
history of |SyncSpec| implied by Lemma~\ref{lem:TwoStepDelayedLinSpec}.  The
fact that $h$ is not synchronisation linearisable means that $h$ and~$h_s'$
are not synchronisation compatible. ............ Actually, there are lots of
different ways that might happen, so it's not clear this discussion is going
to be useful. 


\framebox{Generalisations}: non-binary and/or non-homogeneous. 

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% \framebox{Previous version, not quite right}

%% However, if the delays are sufficiently small, then the
%% log history agrees with the invocation history on the |op|\s1 and~|op|\s2
%% events.  We show that in this case that log history is not linearisable.
%% %
%% \begin{lemma}
%% Consider an invocation history~$h$ that is not synchronisation linearisable
%% with respect to $SyncSpec$.  Let $h_l$ be a corresponding log~history that
%% agrees with~$h$ on |op|\s1 and~|op|\s2 events.  Then $h_l$ is not linearisable
%% with respect to |Two|\-|Step|\-|Delayed|\-|LinSpec|.
%% \end{lemma}

%% %%%%%

%% \begin{proof}
%% We prove the contrapositive: we suppose that $h_l$ is linearisable with
%% respect to |Two|\-|Step|\-|Delayed|\-|LinSpec|, and show that $h$ is
%% synchronisation linearisable with respect to |SyncSpec|. 

%% From the assumption that $h_l$ is linearisable, there is a legal history~$h_s$
%% of |Two|\-|Step|\-|Delayed|\-|LinSpec| such that $h_l$ and~$h_s$ are
%% compatible.  Consider the interleaving of $h_l$ and~$h_s$.  Since $h_l$
%% and~$h$ agree on the |op|\s1 and~|op|\s2 events, we can build a corresponding
%% interleaving of $h$ and~$h_s$: so each $\sm{op}_1^{i_1}(t,x_1)\::()$ event is
%% between events $\call^{inv}.\sm{op}_1^{i_1}(x_1)$ and
%% $\return^{inv}.\sm{op}_1^{i_1} \:: y_1$, where $t$ is the identity of the
%% thread making the call; and each $\sm{op}_2^{i_2}(x_2)\::y_2$ event is
%% between events $\call^{inv}.\sm{op}_2^{i_2}(x_2)$ and
%% $\return^{inv}.\sm{op}_2^{i_2} \:: y_2$.   

%% In |Two|\-|Step|\-|Delayed|\-|LinSpec|, each invocation of $\op_2(x_2)$ calls
%% $\sm{SyncSpec.sync}(x_1, x_2)$, obtaining result $(y_1, y_2)$, and
%% returns~$y_2$.  Here~$x_1$ is taken from |state|, and matches the parameter
%% of the previous  $\sm{op}_1^{i_1}(t,x_1)\::()$ event in~$h_s$.

%% %
%% We build a history $h_s'$ of~|SyncSpec| by replacing each event
%% $\sm{op}_2^{i_2}(x_2)\::y_2$ from~$h_s$ by $\sm{sync}^{i_1,i_2}(x_1, x_2) \::
%% (y_1,y_2)$ for the corresponding~$x_1$ and~$i_1$.  This is a legal history
%% of~|SyncSpec| by construction.  Likewise, each such event is between the
%% corresponding events $\call^{inv}.\sm{op}_2^{i_2}(x_2)$ and
%% $\return^{inv}.\sm{op}_2^{i_2} \:: y_2$, by construction.  
%% \end{proof}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
